---
title: "GSE00910 - Differentially Expressed Genes - Liver Disease"
author: "Paula Unger Avila"
date: "12/6/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About data:GSE99010
Title: 'The synergestic effects of western diet and carbon tetrachloride (CCl4) to develop non-alcoholic steatohepatitis (NASH) and liver tumor in mice' 

Experiment: 'We created western diet-fed and CCl4-treated mouse model. they showed characteristic histological features of NASH such as fibrosis, ballooning and steatosis at 12 weeks and liver cancers developed at 24 weeks'

Each sample contains pooled tissue from 4 mice. 

Disease models: 
Datasets from two time points (12 weeks and 24 weeks )
1) Western Diet 
2) CCI4 

-> will compare two samples to two samples (8 mice)

```{r count table, echo= FALSE}

library(DESeq2)

#setwd()
GSE_ID <- 'GSE99010'
temp <- read.csv('GSE99010_count_table', sep = ' ')
annotation <- read.csv('/Users/paulaungera/Desktop/Project Module Beyer /GSE99010/SraRunTable.txt')

# BEWARE (!!!)  it is very important to check manually that the columns of the count matrix correspond to the rows of the sample information table.

#to have an overview for PCA of all samples together
counts <- as.matrix(temp[,-1])
rownames(counts) <- temp$gene_symbol
annotation[annotation$tissue == 'hepatocellular carcinoma', 'tissue'] <- c('HCC','HCC')
ds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData = annotation,
  design = ~ diet + ccl4_treatment )
levels(ds$diet)
levels(ds$ccl4_treatment)
ds$ccl4_treatment <- relevel(ds$ccl4_treatment, 'Oil')


#Two groups: 
#two last samples are from carcinoma tissue, wont be included
#western diet
counts_wd <-as.matrix(temp[,c(2,4,6,8)])
rownames(counts_wd) <- temp$gene_symbol 
annotation_wd <- annotation[c(1,3,5,7),]

#CCI4
counts_cci4 <-as.matrix(temp[,c(2,3,6,7)])
rownames(counts_cci4) <- temp$gene_symbol 
annotation_cci4 <- annotation[c(1,2,5,6),]

# create a DESeqDataSet 
ds_wd <- DESeqDataSetFromMatrix(
  countData = counts_wd,
  colData = annotation_wd,
  design = ~ diet )
levels(ds_wd$diet)
#check: control is first level

ds_cci4 <- DESeqDataSetFromMatrix(
  countData = counts_cci4,
  colData = annotation_cci4,
  design = ~ ccl4_treatment)
ds_cci4$ccl4_treatment <- relevel(ds_cci4$ccl4_treatment, 'Oil')
levels(ds_cci4$ccl4_treatment)

# remove rows with zero counts
ds_wd <- ds_wd[rowSums(counts(ds_wd)) > 1,]
ds_cci4 <- ds_cci4 [rowSums(counts(ds_cci4 )) > 1,]

#vs transformed counts 
vsd_wd <- vst(ds_wd, blind = T)
vsd_cci4 <- vst(ds_cci4, blind = T)

vsd <- vst(ds,blind = T)

#Euclidean distance between samples (they use transformed data)
sampleDists_wd <- dist(t(assay(vsd_wd)))
sampleDists_cci4 <- dist(t(assay(vsd_cci4)))
sampleDists <- dist(t(assay(vsd)))
```

## Exploring Data

1. Heatmap of sample-to-sample distances using the variance stabilizing transformed values 

```{r heatmap, echo=FALSE}
library("pheatmap")
library("RColorBrewer")

sampleDistMatrix <- as.matrix( sampleDists_wd )
rownames(sampleDistMatrix) <- paste(vsd_wd$diet,"-",vsd_wd$diet_duration)
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists_wd,
         clustering_distance_cols = sampleDists_wd,
         col = colors)


sampleDistMatrix <- as.matrix( sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$diet,"-",vsd$diet_duration, "-", vsd$ccl4_treatment, "-", vsd$tissue)
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)
```



```{r heatmap2, echo=FALSE}

sampleDistMatrix <- as.matrix( sampleDists_cci4 )
rownames(sampleDistMatrix) <- paste(vsd_cci4$ccl4_treatment,"-",vsd_cci4$diet_duration)
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists_cci4,
         clustering_distance_cols = sampleDists_cci4,
         col = colors)
```


3. PCA plot

```{r PCA, echo = FALSE}
plotPCA(vsd_wd, intgroup = c('diet','diet_duration'))
plotPCA(vsd, intgroup = c('diet', 'ccl4_treatment', 'tissue'))

```

```{r PCA, echo = FALSE}
plotPCA(vsd_cci4, intgroup = c('ccl4_treatment', 'diet', 'diet_duration'))

```

## Differential Expression Analysis 

```{r differential expression analysis, echo=TRUE}

ds_wd <- DESeq(ds_wd)
ds_cci4 <- DESeq(ds_cci4)

print(summary(results(ds_wd)))
print(summary(results(ds_cci4)))

res_wd <- results(ds_wd)
res_cci4 <- results(ds_cci4)

```

## Output for liver disease score 

Desired output is a table containing a subset of 200 genes with lowest p-values (not adjusted) and the following columns: (ENSEMBLE Gene IDs), Gene Symbol, log2FoldChange, p-value and p-adjusted. 

In case the gene symbols are misssing in the available count table, the Bioconductor package AnnotationDbi can be used to map the IDs to alternative gene names. 


```{r output, echo=FALSE}
# experimental design: ~ diet_duration + diet
# and ~ diet_duration + ccl4_treatment
res_wd <- res_wd[order(res_wd$pvalue),]
res_cci4 <- res_cci4[order(res_cci4$pvalue),]

output_wd <- data.frame('Gene Symbol' = rownames(res_wd), 
                    'log2FoldChange' = res_wd$log2FoldChange,
                    'pvalue' = res_wd$pvalue,
                    'padj' = res_wd$padj)
write.csv(output_wd, file = paste0(GSE_ID, '_WesternDiet_deg.csv'), 
          row.names = FALSE)



output_cci4 <- data.frame('Gene Symbol' = rownames(res_cci4), 
                    'log2FoldChange' = res_cci4$log2FoldChange,
                    'pvalue' = res_cci4$pvalue,
                    'padj' = res_cci4$padj)
write.csv(output_cci4, file = paste0(GSE_ID, '_CCl4_deg.csv'), 
          row.names = FALSE)

```

Top of the list of differentially expressed genes ordered by lowest p-value:

```{r output, echo = TRUE}
head(res_wd)
head(res_cci4)


```