---
title: "GSE119953 - Differentially Expressed Genes - Liver Disease"
author: "Paula Unger Avila"
date: "12/6/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About data: GSE119953
Title: 'Animal model of acute-on-chronic alcoholic liver injury [DDC study]' 

Experiment design: 
'Adult male C57BL6/J mice were treated with 3,5-diethoxycarbonyl-1,4-dihydrocolidine (DDC) containing diet (0.05% w/w) to induce chronic liver fibrosis. Following DDC-induced fibrogenesis, ethyl alcohol (EtOH) (up to 27 g/ kg/day, up to 28 days) was administered continuously to mice via a gastric feeding tube (Tsukamoto-Frenchmodel of alcoholic liver disease).'
(!!!) maybe ask Thomas about this disease model, not sure if I understand it right

Disease models: 
DDC to induce chronic liver fibrosis (and Tsukamoto-Frenchmodel of alcoholic liver disease ?)

Samples that will be used for:
two time-points: 0D and 28D after DDC application 
1) DDC vs Control (BD)
GSM3389290	#1 Control 0D
GSM3389291	#2 Control 0D
GSM3389292	#3 Control 0D
GSM3389293	#4 Control 0D
GSM3389294	#5 DDC 0D
GSM3389295	#6 DDC 0D
GSM3389296	#7 DDC 0D
GSM3389297	#8 DDC 0D

GSM3389298	#9 Control 28D
GSM3389299	#10 Control 28D
GSM3389300	#11 Control 28D
GSM3389301	#12 Control 28D
GSM3389302	#13 Control 28D

GSM3389313	#24 DDC+Cont ??28 days???
GSM3389314	#25 DDC+Cont
GSM3389315	#26 DDC+Cont
GSM3389316	#28 DDC+Cont
GSM3389317	#29 DDC+Cont

2) High Fat Diet vs. Control (BD)
          
GSM3389290	#1 Control 0D
GSM3389291	#2 Control 0D
GSM3389292	#3 Control 0D
GSM3389293	#4 Control 0D
GSM3389298	#9 Control 28D
GSM3389299	#10 Control 28D
GSM3389300	#11 Control 28D
GSM3389301	#12 Control 28D
GSM3389302	#13 Control 28D
GSM3389303	#14 HFD
GSM3389304	#15 HFD
GSM3389305	#16 HFD
GSM3389306	#17 HFD
GSM3389307	#18 HFD
GSM3389308	#19 HFD


BD = basal diet (I suspect. It's another way of saying normal/chow )

```{r count table}
library(DESeq2)

#setwd()
GSE_ID <- 'GSE119953'
temp <- read.csv('GSE119953_count_table', sep = ' ')
annotation <- read.csv('SraRunTable.txt')

## keep all samples to visualize everything together for later steps 
annotation_all <- annotation 
annotation_all$time <- c(rep('0 days', 8), rep('28 days', 32))
annotation_all$treatment_summed <- paste(annotation_all$treatment_1, annotation_all$treatment_2, annotation_all$time) 
counts_all <- as.matrix(temp[,-1])
rownames(counts_all) <- temp$gene_symbol
colnames(counts_all) <- annotation_all$Run

ds_all <- DESeqDataSetFromMatrix(
  countData = counts_all,
  colData = annotation_all,
  design = ~ treatment_summed)
levels(ds_all$treatment_summed)
ds_all <- ds_all[rowSums(counts(ds_all)) > 1,]
vsd_all <- vst(ds_all, blind = T)
sampleDists_all  <- dist(t(assay(vsd_all)))

# BEWARE (!!!)  it is very important to check manually that the columns of the count matrix correspond to the rows of the sample information table.

annotation <- annotation[c(1:28),]
counts <-as.matrix(temp[,c(2:29)])
rownames(counts) <- temp$gene_symbol 
colnames(counts) <- annotation$Run
#my own annotations of how I understand the data from GEO page + RUN annotations
annotation$treatment_sum <- c('control','control', 'control', 'control',
                              'DDC','DDC','DDC','DDC',
                              'control','control','control','control','control',
                              'HFD','HFD','HFD','HFD','HFD','HFD',
                              'HFD EtOH', 'HFD EtOH','HFD EtOH','HFD EtOH', 
                              'DDC','DDC','DDC','DDC','DDC')
annotation$days <- c('0','0','0','0','0','0','0','0','28','28','28','28','28','28','28','28','28','28','28','28','28','28','28','28','28','28','28','28')
remove(temp)

#subset for two disease models 
# 1) DDC
counts_DDC <- counts[,annotation[annotation$treatment_sum=='control' | annotation$treatment_sum== 'DDC', 'Run' ]]

annotation_DDC <- annotation[annotation$treatment_sum=='control' | annotation$treatment_sum== 'DDC',]

# 2) HFD
counts_HFD <- counts[,annotation[annotation$treatment_sum=='control' | annotation$treatment_sum== 'HFD', 'Run' ]]

annotation_HFD <- annotation[annotation$treatment_sum=='control' | annotation$treatment_sum== 'HFD',]

# create a DESeqDataSet 
ds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData = annotation,
  design = ~ treatment_sum)
levels(ds$treatment_sum)

#1) # create a DESeqDataSet 
ds_DDC <- DESeqDataSetFromMatrix(
  countData = counts_DDC,
  colData = annotation_DDC,
  design = ~ treatment_sum)
levels(ds_DDC$treatment_sum)

#2) # create a DESeqDataSet 
ds_HFD <- DESeqDataSetFromMatrix(
  countData = counts_HFD,
  colData = annotation_HFD,
  design = ~ treatment_sum)
levels(ds_HFD$treatment_sum)


# remove rows with zero counts
ds <- ds[rowSums(counts(ds)) > 1,]
ds_DDC <- ds_DDC[rowSums(counts(ds_DDC)) >1,]
ds_HFD <- ds_HFD[rowSums(counts(ds_HFD)) >1,]

#vs transformed counts 
vsd <- vst(ds, blind = T)
vsd_DDC <- vst(ds_DDC, blind = T)
vsd_HFD <- vst(ds_HFD, blind = T)

#Euclidean distance between samples (they use transformed data)
sampleDists  <- dist(t(assay(vsd)))
sampleDists_DDC <- dist(t(assay(vsd_DDC)))
```

## Exploring Data

1. Heatmap of sample-to-sample distances using the variance stabilizing transformed values 

```{r heatmap, echo=FALSE}
library("pheatmap")
library("RColorBrewer")

sampleDistMatrix <- as.matrix( sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$treatment_sum,"-",vsd$days)
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)
```


3. PCA plot
-> consider removing two outliers??? 

```{r PCA, echo = FALSE}
plotPCA(vsd, intgroup = c('treatment_sum','days'))

```

```{r PCA, echo = FALSE}
plotPCA(vsd_all, intgroup = c('treatment_summed'))

```

# Heatmap and PCA plot only DDC data

```{r heatmap subset, echo=FALSE}
sampleDistMatrix <- as.matrix( sampleDists_DDS)
rownames(sampleDistMatrix) <- paste(vsd_DDC$treatment_sum,"-",vsd_DDC$days)
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists_DDS,
         clustering_distance_cols = sampleDists_DDS,
         col = colors)
```
```{r PCA subset, echo = FALSE}
plotPCA(vsd_DDC, intgroup = c('treatment_sum','days'))
plotPCA(vsd_HFD, intgroup = c('treatment_sum','days'))

```
```{r heatmap subset, echo=FALSE}
sampleDistMatrix <- as.matrix( sampleDists_all)
rownames(sampleDistMatrix) <- paste(vsd_all$treatment_summed)
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists_all,
         clustering_distance_cols = sampleDists_all,
         col = colors)
```

## Differential Expression Analysis 

```{r differential expression analysis, echo=FALSE}

ds_DDC <- DESeq(ds_DDC)
ds_HFD <- DESeq(ds_HFD)
print(summary(results(ds_DDC)))
print(summary(results(ds_HFD)))

res_DDC <- results(ds_DDC)
res_HFD <- results(ds_HFD)

```

## Output for liver disease score 

Desired output is a table containing a subset of 200 genes with lowest p-values (not adjusted) and the following columns: (ENSEMBLE Gene IDs), Gene Symbol, log2FoldChange, p-value and p-adjusted. 

In case the gene symbols are missing in the available count table, the Bioconductor package AnnotationDbi can be used to map the IDs to alternative gene names. 


```{r output, echo=FALSE}
# experimental design: ~ agent
res_DDC <- res_DDC[order(res_DDC$pvalue),]
res_HFD <- res_HFD[order(res_HFD$pvalue),]

write.csv(data.frame('Gene Symbol' = rownames(res_DDC), 
                    'log2FoldChange' = res_DDC$log2FoldChange,
                    'pvalue' = res_DDC$pvalue,
                    'padj' = res_DDC$padj), 
          file = paste(GSE_ID,'_DDCn_deg.csv'), 
          row.names = FALSE)

write.csv(data.frame('Gene Symbol' = rownames(res_HFD), 
                    'log2FoldChange' = res_HFD$log2FoldChange,
                    'pvalue' = res_HFD$pvalue,
                    'padj' = res_HFD$padj), 
          file = paste(GSE_ID,'_HFD_deg.csv'), 
          row.names = FALSE)
head(res_DDC)
head(res_HFD)
#remove(list = ls())

```
