---
title: "GSE119441 - Differentially Expressed Genes - Liver Disease"
author: "Paula Unger Avila"
date: "12/6/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About data: GSE119441
Title: 'The effects of perfluorooctanoate on high fat diet induced non-alcoholic fatty liver disease in mice' 

Experiment design: 
'Two groups of mice were fed with either a control or high-fat diet for 16 weeks to induce normal or steatotic livers, respectively. Subsequently, mice from each diet group were treated with PFOA or vehicle for another 8 weeks. Each mouse was considered a biological replicate. A total of 4 groups with 4 mice in each group were used.'


Disease models: two disease models 

1) normal diet + vehicle vs HFD + vehicle 
2) normal diet + vehicle vs normal diet + PFOA

4 samples per condition 



```{r count table}

library(DESeq2)

#setwd()
GSE_ID <- 'GSE119441'
temp <- read.delim('GSE119441_genecounts.txt', sep = '\t')
annotation <- read.csv('SraRunTable.txt')

# BEWARE (!!!)  it is very important to check manually that the columns of the count matrix correspond to the rows of the sample information table.
counts <- as.matrix(temp[,c(2:17)])
colnames(counts) <- annotation$Run
rownames(counts) <- temp$id

counts_HFD <- counts[,c(1:8)] 
ann_HFD <- annotation[c(1:8),]

counts_PFOA <- counts[,c(1,2,3,4,13,14,15,16)]
ann_PFOA <- annotation[c(1,2,3,4,13,14,15,16),]


# create a DESeqDataSet 
ds_HFD <- DESeqDataSetFromMatrix(
  countData = counts_HFD,
  colData = ann_HFD,
  design = ~ treatment )

levels(ds_HFD$treatment)

ds_PFOA <- DESeqDataSetFromMatrix(
  countData = counts_PFOA,
  colData = ann_PFOA,
  design = ~ treatment )

ds_PFOA$treatment <- relevel(ds_PFOA$treatment, 'CD + Vehicle')
levels(ds_PFOA$treatment)

ds <-  DESeqDataSetFromMatrix(
  countData = counts,
  colData = annotation,
  design = ~ treatment )

ds$treatment <- relevel(ds$treatment, 'CD + Vehicle' )
levels(ds$treatment)
  

# remove rows with zero counts
ds_HFD <- ds_HFD[rowSums(counts(ds_HFD)) > 1,]
ds_PFOA <- ds_PFOA[rowSums(counts(ds_HFD)) > 1,]
ds <- ds[rowSums(counts(ds)) > 1,]
#vs transformed counts 
vsd_hfd <- vst(ds_HFD, blind = T)
vsd_pfoa <- vst(ds_PFOA, blind = T)
vsd <- vst(ds, blind = T)
#Euclidean distance between samples (they use transformed data)
sampleDists_hfd  <- dist(t(assay(vsd_hfd)))
sampleDists_pfoa  <- dist(t(assay(vsd_pfoa)))
sampleDists <- dist(t(assay(vsd)))
```

## Exploring Data

1. Heatmap of sample-to-sample distances using the variance stabilizing transformed values 

```{r heatmap, echo=FALSE}
library("pheatmap")
library("RColorBrewer")

sampleDistMatrix <- as.matrix( sampleDists)
rownames(sampleDistMatrix) <- vsd$treatment
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)
```
```{r heatmap, echo=FALSE}

sampleDistMatrix <- as.matrix( sampleDists_pfoa)
rownames(sampleDistMatrix) <- vsd_pfoa$treatment
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists_pfoa,
         clustering_distance_cols = sampleDists_pfoa,
         col = colors)
```

3. PCA plot

```{r PCA, echo = FALSE}
plotPCA(vsd_hfd, intgroup = 'treatment')
plotPCA(vsd, intgroup = 'treatment')
```

```{r PCA, echo = FALSE}
plotPCA(vsd_pfoa, intgroup = 'treatment')

```


## Differential Expression Analysis 

```{r differential expression analysis, echo=TRUE}

ds_HFD <- DESeq(ds_HFD)
ds_PFOA<- DESeq(ds_PFOA)

print(summary(results(ds_HFD)))
print(summary(results(ds_PFOA)))

res_HFD <- results(ds_HFD)
res_PFOA <- results(ds_PFOA)

```

## Output for liver disease score 

Desired output is a table containing a subset of 200 genes with lowest p-values (not adjusted) and the following columns: (ENSEMBLE Gene IDs), Gene Symbol, log2FoldChange, p-value and p-adjusted. 

In case the gene symbols are missing in the available count table, the Bioconductor package AnnotationDbi can be used to map the IDs to alternative gene names. 


```{r output, echo=FALSE}
# experimental design: ~ agent
res_HFD <- res_HFD[order(res_HFD$pvalue),]
res_PFOA <- res_PFOA[order(res_PFOA$pvalue),]

output_HFD <- data.frame('Gene Symbol' = rownames(res_HFD), 
                    'log2FoldChange' = res_HFD$log2FoldChange,
                    'pvalue' = res_HFD$pvalue,
                    'padj' = res_HFD$padj)

output_PFOA <- data.frame('Gene Symbol' = rownames(res_PFOA), 
                    'log2FoldChange' = res_PFOA$log2FoldChange,
                    'pvalue' = res_PFOA$pvalue,
                    'padj' = res_PFOA$padj)

write.csv(output_HFD, file = paste(GSE_ID, '_HFD_ensembleid_deg.csv'), 
          row.names = FALSE)
write.csv(output_PFOA, file = paste(GSE_ID, '_PFOA_ensembleid_deg.csv'), 
          row.names = FALSE)

```
